### 2025-07-18 paper/quarto_aaai/paper.tex
methodology adjusted and algorithm added

CEs have also been used to improve models in the natural language
processing domain. For example, \citet{wu2021polyjuice2} propose
\emph{Polyjuice}, a general-purpose CE generator for language models and
demonstrate that the augmentation of training data with \emph{Polyjuice}
improves robustness in a number of tasks, while
\citet{luu2023counterfactual} introduce the \emph{Counterfactual
Adversarial Training} (CAT) framework that aims to improve
generalization and robustness of language models by generating
counterfactuals for training samples that are subject to high predictive
uncertainty.

@@

This section introduces the Counterfactual Training framework. CT
combines ideas from adversarial training, counterfactual explanations,
and energy-based modelling with the explicit goal of producing models
whose learned representations align with plausible explanations that
further comply with user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which generated CEs comply with the
underlying data-generating process
\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. An AR recommendations must also be actionable, i.e.,
possible to attain by the recipient. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they had been 20 years
older. Ignoring all other features, this would comply with the
definition of plausibility if 40-year-old individuals were in fact more
credit-worthy on average than young adults. But to the applicant, this
CE does not provide actionable recourse, as age is not a (directly)
mutable feature. Counterfactual training aims to improve model
explainability by aligning models with counterfactuals that meet both
desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

\subsection{Proposed Objective}\label{proposed-objective}

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. In broad terms, searching
for CEs using gradient descent entails optimizing some form of an
objective that balances (1) the classification loss
\(\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})\), and (2) one
or more penalty terms \(\lambda_{i}\text{cost}_{i}(\cdot)\). The exact
specification of these penalties induces various properties in the
counterfactual outcomes, and tends to be the key feature that
distinguishes various gradient-based ``generators'' or ``explainers'' in
the literature \citep{altmeyer2023explaining}, including all generators
used in our experiments. We refer the reader to the supplementary
appendix for details.

CT adopts gradient-based CE search during training to generate
on-the-fly model explanations \(\mathbf{x}^\prime\) for training
samples. We use the term \emph{nascent} to denote counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) that are not yet valid where \(t\)
indicates the last iteration before the label is flipped. We store and
use these interim counterfactuals as adversarial examples. Conversely,
we consider counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature}
explanations if they have either exhausted all \(T\) iterations or
converged in terms reaching a pre-specified threshold, \(\tau\), for the
predicted probability of the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models:
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing the contrastive divergence, \(\text{div}(\cdot)\), between
mature counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and the adversarial loss, \(\text{advloss}(.)\), wrt.
nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \citep{du2019implicit}. The trade-off between the
components are governed through \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

>>>>

CEs have also been used to improve models in the natural language
processing domain. For example, \citet{wu2021polyjuice} propose
\emph{Polyjuice}, a general-purpose CE generator for language models and
demonstrate that the augmentation of training data with \emph{Polyjuice}
improves robustness in a number of tasks, while
\citet{luu2023counterfactual} introduce the \emph{Counterfactual
Adversarial Training} (CAT) framework that aims to improve
generalization and robustness of language models by generating
counterfactuals for training samples that are subject to high predictive
uncertainty.

@@

This section introduces the counterfactual training framework. As
mentioned above, CT applies ideas from contrastive and robust learning
to counterfactual explanations in order to produce models whose learned
representations align with plausible explanations that further comply
with user-defined actionability constraints.

Counterfactual explanations are typically generated by solving
variations of the following optimization problem,
\begin{equation}\phantomsection\label{eq-general}{
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
}\end{equation} where
\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
\(D\) features and \(\mathbf{y}^+\) denotes some target class. The
\(\text{yloss}(\cdot)\) function quantifies the discrepancy between
current model predictions for \(\mathbf{x}^{\prime}\) and the target
class (a conventional choice is cross-entropy). Finally, we use
\(\text{reg}(\cdot)\) to denote any form of regularization used to
induce certain properties on the counterfactual. In their seminal paper,
\citet{wachter2017counterfactual} propose regularizing the distance
between counterfactuals and their original factual values, to ensure
that individuals seeking recourse through CE face minimal costs in terms
of feature changes. Different variations of Equation~\ref{eq-general}
have been proposed in the literature to address various other desiderata
including those already introduced above (faithfulness, plausibility and
actionability). Like \citet{wachter2017counterfactual}, most of these
approaches rely on gradient descent to optimize
Equation~\ref{eq-general}. For more details on various approaches tested
in this work, we refer the reader to the supplementary appendix. In the
following, we describe in detail how counterfactuals are generated and
used in CT.

\subsection{Proposed Training
Objective}\label{proposed-training-objective}

The principal goal of CT is to improve model explainability by aligning
models with faithful explanations that are plausible and actionable.
Formally, we define explainability as follows:

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. CT adopts gradient-based
CE search during training to generate on-the-fly model explanations
\(\mathbf{x}^\prime\) for training samples. We use the term
\emph{nascent} to denote interim counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
explain below, these nascent counterfactuals can be stored and
repurposed as adversarial examples. Conversely, we consider
counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
they have either exhausted all \(T\) iterations or converged in terms
reaching a pre-specified threshold, \(\tau\), for the predicted
probability of the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models:
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing the contrastive divergence, \(\text{div}(\cdot)\), between
mature counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and the adversarial loss, \(\text{advloss}(.)\), wrt.
nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \citep{du2019implicit}. The trade-off between the
components are governed through \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\). The full training
regime is sketched out in Algorithm \ref{alg-experiment}.

\begin{algorithm}[h]
  \caption{Counterfactual Training}
    \label{alg-experiment}
    \begin{algorithmic}[1]
    \REQUIRE Training dataset $\mathcal{D}$, initial model $\mathbf{M}_{\theta_0}$
    \WHILE{not converged}
        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
        \FOR{$t = 1$ to $T$}
            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
        \ENDFOR
        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
    \ENDWHILE
    \RETURN $\mathbf{M}_\theta$
    \end{algorithmic}
\end{algorithm}

### 2025-07-18 paper/sections/experiments.qmd
methodology adjusted and algorithm added

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). They are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix. Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2. Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].

>>>>

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). They are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix. Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give. Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].

### 2025-07-18 paper/sections/method.qmd
methodology adjusted and algorithm added

This section introduces the Counterfactual Training framework. CT combines ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of producing models whose learned representations align with plausible explanations that further comply with user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which generated CEs comply with the underlying data-generating process [@altmeyer2024faithful;@guidotti2022counterfactual;@poyiadzi2020face]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models. An AR recommendations must also be actionable, i.e., possible to attain by the recipient. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they had been 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. Counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability. Formally, we define explainability as follows:

@@

## Proposed Objective

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$. The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" in the literature [@altmeyer2023explaining], including all generators used in our experiments. We refer the reader to the supplementary appendix for details.

CT adopts gradient-based CE search during training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for training samples. We use the term *nascent* to denote counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that are not yet valid where $t$ indicates the last iteration before the label is flipped. We store and use these interim counterfactuals as adversarial examples. Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged in terms reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

>>>>

This section introduces the counterfactual training framework. As mentioned above, CT applies ideas from contrastive and robust learning to counterfactual explanations in order to produce models whose learned representations align with plausible explanations that further comply with user-defined actionability constraints.

Counterfactual explanations are typically generated by solving variations of the following optimization problem,
$$
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
$$ {#eq-general}
where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+$ denotes some target class. The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy). Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual. In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values, to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes. Different variations of @eq-general have been proposed in the literature to address various other desiderata including those already introduced above (faithfulness, plausibility and actionability). Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general. For more details on various approaches tested in this work, we refer the reader to the supplementary appendix. In the following, we describe in detail how counterfactuals are generated and used in CT.

## Proposed Training Objective

The principal goal of CT is to improve model explainability by aligning models with faithful explanations that are plausible and actionable. Formally, we define explainability as follows:

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. CT adopts gradient-based CE search during training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for training samples. We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that have not yet converged. As we explain below, these nascent counterfactuals can be stored and repurposed as adversarial examples. Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged in terms reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$. The full training regime is sketched out in Algorithm \ref{alg-experiment}.

\begin{algorithm}[h]
  \caption{Counterfactual Training}
    \label{alg-experiment}
    \begin{algorithmic}[1]
    \REQUIRE Training dataset $\mathcal{D}$, initial model $\mathbf{M}_{\theta_0}$
    \WHILE{not converged}
        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
        \FOR{$t = 1$ to $T$}
            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
        \ENDFOR
        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
    \ENDWHILE
    \RETURN $\mathbf{M}_\theta$
    \end{algorithmic}
\end{algorithm}

### 2025-07-18 paper/sections/related.qmd
methodology adjusted and algorithm added

CEs have also been used to improve models in the natural language processing domain. For example, @wu2021polyjuice2 propose *Polyjuice*, a general-purpose CE generator for language models and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks, while @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.

>>>>

CEs have also been used to improve models in the natural language processing domain. For example, @wu2021polyjuice propose *Polyjuice*, a general-purpose CE generator for language models and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks, while @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
